Tidy data structure to support exploration and modeling of temporal-context data

### 2018-04-23 main.Rmd
opening paragraph in intro

---
title: "Tidy data structures to support exploration and modeling of temporal-context data"
author:
  - familyname: Wang
    othernames: Earo
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : earo.wang@monash.edu
    correspondingauthor: true
  - familyname: Cook
    othernames: Dianne
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : dicook@monash.edu
  - familyname: Hyndman
    othernames: Rob J
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : rob.hyndman@monash.edu
abstract: 'Temporal-context data is often rich with information and time formats, for example multiple units, fine time resolutions, heterogeneous data types, nested and crossed structures and etc. This work presents a cohesive and conceptual framework for organizing and manipulating temporal data, which in turn flows into visualization and modelling routines. Tidy data principles are applied and extended to temporal data: (1) mapping the semantics of a dataset into its physical layout, (2) an explicitly declared variable representing time index, (3) single or multiple identifiers separating multiple units over time, which are referred to as a "key". A syntax is proposed to describe nested or crossed data structures, which employ the "key". The tidy temporal data structure most naturally supports thinking of operations on the data as building blocks, forming part of a "data pipeline". A sound pipeline practice facilitates a transparent and human-readable workflow for analyzing temporal data. Applications are included to illustrate the tidy temporal data, data pipeline ideas and usage. A new representation of tidy temporal data along with the supporting methods has been implemented in the R package **tsibble**.'
keywords: "temporal context, time series, data structure, R"
wpnumber: no/yr
jelcodes: C10,C14,C22
blind: false
cover: true
toc: false
bibliography: references.bib
biblio-style: authoryear-comp
output:
  MonashEBSTemplates::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
    keep_tex: no
    number_sections: yes
    citation_package: biblatex
---

@@

Data measured or collected at different time points are referred to as temporal-context data or time series. This dataset arrives in a rectangular form, which features:

* heterogeneous data columns, such as numbers, characters and timestamps;
* multiple measured variables including departure/arrival delay in minutes, flight time, distance between origin and destination, and etc.;
* low time resolutions, for example scheduled departure time in minutes;
* multi-levels of grouping like airports nested within cities, and cities further nested within states.

However, there is lack of data management tools for time series involving those characteristics in most statistical software, for example data wrangling, visualisation, analysing and forecasting in time-based context. It becomes more challenging in handling time series as increased data complications. The bottleneck that makes time series data frustrating to work with is the conventional data structure that a vector or matrix underlies.

@@

@@

# Tidy temporal-context data

>>>>

---
title: "Tidy data structure to support exploration and modeling of temporal-context data"
author:
  - familyname: Wang
    othernames: Earo
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : earo.wang@monash.edu
    correspondingauthor: true
  - familyname: Cook
    othernames: Dianne
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : dicook@monash.edu
  - familyname: Hyndman
    othernames: Rob J
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : rob.hyndman@monash.edu
abstract: 'Temporal-context data is often rich with information and time formats, for example multiple observational units, different time lengths, heterogeneous data types, nested and crossed factors and etc. This work presents a cohesive and conceptual framework for organizing and manipulating temporal data, which in turn flows into visualization and modelling routines. Tidy data principles are applied and extended to temporal data: (1) mapping the semantics of a dataset into its physical layout, (2) an explicitly declared index variable representing time, (3) a "key" comprised of single or multiple variables to uniquely identify units over time, using a syntatical and user-oriented approach in which it imposes nested or crossed structures on the data. This tidy data representation most naturally supports thinking of operations on the data as building blocks, forming part of a "data pipeline" in time-based context. A sound pipeline practice facilitates a transparent and human-readable workflow for analyzing temporal data. Applications are included to illustrate tidy temporal data structure, data pipeline ideas and usage. The tidy temporal data along with its supporting methods has been implemented in the R package **tsibble**.'
keywords: "temporal context, time series, data structure, R"
wpnumber: no/yr
jelcodes: C10,C14,C22
blind: false
cover: true
toc: false
bibliography: references.bib
biblio-style: authoryear-comp
output:
  MonashEBSTemplates::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
    keep_tex: no
    number_sections: yes
    citation_package: biblatex
---

@@

Temporal-context data consist of observational units indexed at different time points $X_{jt}$, where the $j$\textsuperscript{th} unit takes measurements of $X$ on over time $t$, for $j = 1, \cdots, N$ and $1 \le t \le T$. Time primarily forms the contextual basis of temporal data, but it could arrive in many possible formats. For example, data recorded at fine time resolutions (hours, minutes, and seconds) are typically associated with different time zones and daylight savings. Temporal data often carries with rich information other than the time: multiple observational units of different time lengths, multiple and heterogeneous measured variables, multiple grouping variables exhibiting nested or crossed structures, linking to other data tables, and etc.

@@

# Data semantics

@@
